### 2014-05-08 ApTranscriptome_TR.Rmd
Adding analysis of temperature expression turned on/off.
Some updates to knit nicely.

Thermal reactionome of the common ant species *Aphaenogaster*
================================================================

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)

@@

# knitr options
opts_chunk$set(cache=TRUE)

@@

```{r}
system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
```

@@

```{r sailfish, eval=TRUE, echo=TRUE, cache=TRUE}
# directory containing trimmed reads
readdir <- "data/ind_files/" 
# sailfish index directory
sfindex <- "results/trinity-full/sailfish-index-Trinity-cap3-uclust"
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"

@@

# define function for `ddply` to calculate overall model P value
modpFunc <- function(foo) {
  # fit lm
  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))

  # calculate overall model significance
  f <- summary(lmout)$fstatistic
  pval <- unname(pf(f[1],f[2],f[3],lower.tail=F))
  attributes(pval) <- NULL

  # return pvalue
  c(pval = pval) 
}

RxNpval <- ddply(TPM.dt.sub, .(Transcript), .progress="text", .inform="TRUE", modpFunc)
```

@@

```{r padjust}
RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
# Plot FDR values against initial pvalues
png(file=paste(resultsdir, "pval_hist.png"))
par(mfrow = c(2,1))
hist(RxNpval$pval)
hist(RxNpval$padj)
dev.off()

@@

```{r lmFunc}
# first, define function to use
lmFunc <- function(bar) {
  # fit lm
  lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))  
  # stepAIC to drop non-significant terms
  f.lmout <- stepAIC(lmout)
  # return final model
  return(f.lmout) 
}

# apply function to all responsive transcripts
RxNlmAIC <- dlply(sig.TPM.dt.sub, .(Transcript), .progress="text", lmFunc)
```

@@

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts (of 105,536 total) at 5% FDR with expression that depends on colony, temperature or their interaction.")
```

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.lms, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.lms))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.lms))

@@

Examine transcripts by temperature response. Compare between colonies.

```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])
# reorder for plotting
tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]

png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
mosaicplot(tt2, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
dev.off()

pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")

# Pearson Chi-square test
chisq.test(Ap.type.table)
```

@@

```{r optimum_expression_comparison, cache=TRUE, eval=FALSE}
# list of transcripts that are 'high' expressed in A22
A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099")`. As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

@@

```{r max_exp_PDF}
# reshape data
Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "ApNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))

@@

```{r offset, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "text", RxNvarshape)

@@

@@

```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
## merge results into single table
Ap.high.gsea$Type <- "High"
Ap.low.gsea$Type <- "Low"
Ap.bim.gsea$Type <- "Bimodal"
Ap.int.gsea$Type <- "Intermediate"
# combine
Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
# reorder
Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea1)[7] <- "P"
# display
pandoc.table(Ap.gsea1, split.tables = "Inf")
```

@@

# Run GSEA
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
#pandoc.table(A22.high.gsea)
```

@@

# Run GSEA
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
```

@@

# Run GSEA
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
```

@@

# Run GSEA
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
```

@@

# Run GSEA
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
```

@@

# Run GSEA
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
```

@@

# Run GSEA
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
```

@@

# Run GSEA
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
```

Make a table with results.

@@

Ar.high.gsea$Type <- "High"
Ar.low.gsea$Type <- "Low"
Ar.bim.gsea$Type <- "Bimodal"
Ar.int.gsea$Type <- "Intermediate"
Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
Ar.gsea$Colony <- "ApNC"

@@

Ap.setdiff.gsea
```

@@

Organize data

```{r plot_data}
# extract TPM data for thermally-responsive transcripts
resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
setkey(resp.TPM.dt.sub, Transcript)
str(resp.TPM.dt.sub)
length(unique(resp.TPM.dt.sub$Transcript))
# scale transcripts so can compare
resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
```

@@

```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

@@

# Smooth plot, expression against temp, faceted by colony
png("results/GO_stress_High.png")
trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp2)
dev.off()
```

@@

```{r plot_GOstress_low, echo=FALSE, eval=FALSE, cache=TRUE}
low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]

@@

```{r references, results='asis'}
bibliography()
```

>>>>

Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
========================================================================================

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)
library(xtable)
library(methods)

@@

# knitr options
opts_chunk$set(cache=TRUE)
options(width = 80)

@@

```{r mk_sailfish_dir}
system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
```

@@

```{r sailfish_quant, eval=TRUE, echo=TRUE, cache=TRUE}
# directory containing trimmed reads
readdir <- "data/ind_files/" 
# sailfish index directory
sfindex <- "results/trinity-full/sailfish-index-Trinity-cap3-uclust"
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"

@@

# calculate overall P value for 'model' for each transcript
RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
```

@@

```{r padjust}
RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
# Plot FDR values against initial pvalues
par(mfrow = c(2,1))
hist(RxNpval$pval)
hist(RxNpval$padj)

@@

```{r RxNlmAIC, echo=TRUE, eval=TRUE, results='hide'}
# perform model selection for responsive transcripts
RxNlmAIC <- try(dlply(sig.TPM.dt.sub, .(Transcript), lmFunc))
```

@@

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on colony, temperature or their interaction at 5% FDR.")
```

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.lms))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.lms))

@@

Next, I compare the number of thermally-responsive in each response category between the two colonies.

```{r temperature_response}
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

# table
Ap.type.table <- rbind(A22.type.table, Ar.type.table)

# Pearson Chi-square test
chisq.test(Ap.type.table)

# Reorganize for plotting to show overlap among categories between colonies
type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])

# reorder 
tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]

# plot
mosaicplot(tt2, xlab = "AcNC", ylab = "ApVT", main = "Mosaic plot of responsive transcripts")
```

The number of thermally-responsive in each response category differs between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies. For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts. For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.

Interestingly, nearly half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.

```{r temperature_response_table, echo=FALSE, results='asis'}
pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
```

@@

```{r optimum_expression_comparison, cache=TRUE, eval=TRUE}
# list of transcripts that are 'high' expressed in A22
A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

@@

```{r max_exp_PDF}
# reshape data
Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))

@@

```{r offset, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)

@@

## Temperature of gene activation

Thermally-responsive genes could also differ in the temperatures at which they have increased or decreased expression in response to temperature changes. To examine this, I determine the temperature at which each responsive gene has the greated increase or decrease in expression.

```{r responsive_TPM}
# extract TPM data for thermally-responsive transcripts
resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
setkey(resp.TPM.dt.sub, Transcript)
str(resp.TPM.dt.sub)
length(unique(resp.TPM.dt.sub$Transcript))
# scale transcripts so can compare
resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
```

Predict for responsive transcripts

```{r responsive_pred}

predFunc <- function(foo) {
  # Function supplied to `ddply` to predict expression at each temperature
  #
  # Args:
  #  foo: data.frame model is fit to
  #
  #  'model' must be specified in environment  
  #
  # Returns:
  #  data.frame of Transcripts with pval

  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))

  # predict values
  pout <- predict(lmout)
  # back-transform from log-scale
  foo$pTPM <- exp(pout)

  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
  foo$pTPM <- ifelse(foo$pTPM < 0, 0, round(foo$pTPM, 5))
  # plot
  #with(foo, qplot(val, TPM, colour = colony, geom = "smooth"))
  #with(foo, qplot(val, pTPM, colour = colony, geom = "smooth"))

  pred.out.df <- data.frame(Transcript = foo$Transcript, TPM = foo$TPM, val = foo$val, colony = foo$colony, pTPM = foo$pTPM)
}

# apply predFunc to responsive transcripts
testpred <- ddply(resp.TPM.dt.sub[1:220], .(Transcript), .inform="TRUE", predFunc)

# calculate temperature of greatest change
T.on.df <- ddply(testpred, .(Transcript, colony), summarise, 
              T_on <- diff(pTPM))

testdiff[(which(diff(testdiff$pTPM) == max(diff(testdiff$pTPM)))+1), ]

```

### Identification of biological-functions associated with temperature responses

@@

```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
## merge results into single table
Ap.high.gsea$Type <- "High"
Ap.low.gsea$Type <- "Low"
Ap.bim.gsea$Type <- "Bimodal"
Ap.int.gsea$Type <- "Intermediate"
# combine
Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
# reorder
Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea1)[7] <- "P"
# display
Ap.gsea1
```

@@

# Run GSEA
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
#pandoc.table(A22.high.gsea, split.table = Inf)
```

@@

# Run GSEA
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
#pandoc.table(Ar.high.gsea, split.table = Inf)
```

@@

# Run GSEA
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
#pandoc.table(A22.low.gsea, split.table = Inf)
```

@@

# Run GSEA
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
#pandoc.table(Ar.low.gsea, split.table = Inf)
```

@@

# Run GSEA
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
#pandoc.table(A22.bim.gsea, split.table = Inf)
```

@@

# Run GSEA
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
#pandoc.table(Ar.bim.gsea, split.table = Inf)
```

@@

# Run GSEA
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
#pandoc.table(A22.int.gsea, split.table = Inf)
```

@@

# Run GSEA
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
#pandoc.table(Ar.int.gsea, split.table = Inf)
```

Table of overall GSEA results.

@@

Ar.high.gsea$Type <- "High"
Ar.low.gsea$Type <- "Low"
Ar.bim.gsea$Type <- "Bimodal"
Ar.int.gsea$Type <- "Intermediate"
Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
Ar.gsea$Colony <- "AcNC"

@@

print(xtable(Ap.setdiff.gsea, comment = getOption("xtable.comment", FALSE)))
```

@@

@@

```{r plot_GOstress, echo=FALSE, eval=TRUE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

@@

# Smooth plot, expression against temp, faceted by colony
trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp2)
```

@@

```{r plot_GOstress_low, echo=FALSE, eval=TRUE, cache=TRUE}
low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]

@@

```{r references, echo = FALSE, results='asis'}
bibliography()
```
